---
title: "Death Registry Statistics"
subtitle: "Subtitle Here"
author: "Xiangyu Wang"
thanks: "Code and data are available at: https://github.com/GuLanTe1/STA304_2022_A1.git."
date: "03 February 2022"
abstract: "abstract here"
output: bookdown::pdf_document2
License: Open Government License - Toronto
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(knitr)
library(usethis)
```

# Introduction

first paragraph is motivational and broad by expanding abstract

second paragraph is about what was done and what was found.

third is about implications

the remainder of this paper is: section \@ref(data) explains the data. section \@ref(results) covers results.....

# Data

paragraph or two introducing the dataset broadly.

Then show an extract of the dataset (kable)

```{r}

death_registry <- read.csv("death_registry.csv")

death_registry$CIVIC_CENTRE <- sub("ET", "Etobicoke", death_registry$CIVIC_CENTRE)

death_registry$CIVIC_CENTRE <- sub("NY", "North York", death_registry$CIVIC_CENTRE)

death_registry$CIVIC_CENTRE <- sub("SC", "Scarborough", death_registry$CIVIC_CENTRE)

death_registry$CIVIC_CENTRE <- sub("TO", "Toronto ", death_registry$CIVIC_CENTRE)


death_registry |>
  select(CIVIC_CENTRE, DEATH_LICENSES, PLACE_OF_DEATH,  TIME_PERIOD) |>
  slice(1:10) |>
  kable(
    caption = "Preview of Death Registry Statistics",
    col.names = c("Civic centre", "Death licenses", "Place of death", "Time period"),
    align = c("c", "c","c","c","c","c"),
    booktabs = TRUE,
    linesep = ""
  )
```

Paragraph or two more about Table 1

Explaniation about interest of the next graph basbdaskjbdskjadbas.











